{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-10-06 thesis/intro.tex
intro subsubsections
assorted minor prose revisions

\subsection{Motivation}

@@

\subsection{Contribution}

@@

\subsection{Organization}

@@

\begin{itemize}
	\item {\bf Background:} Chapter~\ref{chap:background} will present the requisite background material on concurrent programming, stateless model checking, and the various types of programs targeted by Landslide.
	\item {\bf Landslide:} Chapter~\ref{chap:landslide} explains the design and implementation of Landslide
		%the core of my stateless model checking framework,
		and all the special features it's been equipped with over the years.
		It is the foundation upon which all three contributions above build.
	\item {\bf Quicksand:} Chapter~\ref{chap:quicksand} presents the Iterative Deepening framework which more intelligently chooses which state spaces to test, corresponding to contribution 1 above.
	\item {\bf Education:} Chapter~\ref{chap:education} discusses my evaluation of Landslide
		in CMU's 15-410 class environment using the Pebbles kernel,
		and in the University of Chicago's and Berkeley's OS class environments using the Pintos kernel,
		corresponding to contribution 2 above.
	%\item {\bf Pebbles:} Chapter~\ref{chap:410} discusses my evaluation of Landslide in CMU's 15-410 class environment using the Pebbles kernel, corresponding to contribution 2 above.
	%\item {\bf Pintos:} Chapter~\ref{chap:410} discusses my evaluation of Landslide in the University of Chicago's and Berkeley's OS class environments using the Pintos kernel, corresponding to contribution 2 above.
	\item {\bf Transactional Memory:} Chapter~\ref{chap:tm} presents my extension of Landslide's concurrency model to handle transactional concurrency and the evaluation thereof, corresponding to contribution 3 above.
	\item {\bf Related Work:} Chapter~\ref{chap:relatedwork} honors my neighbours and ancestors in research spirit.
	\item {\bf Future Work:} Chapter~\ref{chap:warpzone} % TODO
	\item {\bf Conclusion:} Chapter~\ref{chap:conclusion} provides some thoughts on the future of the field.
\end{itemize}

@@

Pronoun use will vary between more specific and more ambiguous to convey additional nuance.
The singular ``I'' is associated with my own research contributions,
while the royal ``we'' should be taken to include the reader,
such as when surveying background material or related work,
to which the author and reader share more similar relationships.
The impersonal ``the programmer'' will be referred to as she/her
to highlight her role as the intended user, separate from the underlying research,
%to distinguish her from the thesis audience concerned with the underlying research,
and also to challenge readers' unconscious bias about gender in computer science.
Individual students who participated in the user studies will be given the more inclusive they/them.
Gender-neutral pronouns will also be used on the author themself.

This document is, in a way, only half the work of the thesis, the other half being Landslide's implementation.
While some readers may prefer to be taught in prose and/or mathematical notation how an algorithm works,
others may find that disorienting and wish to see things in a way a compiler would understand.
The beginning of Chapter~\ref{chap:landslide} provides source code links,
and the rest of it serves as a guide to browsing the repository.
Later chapters will often make parenthetical references to specific files and/or functions
therein which implement a feature under discussion.

>>>>

\subsubsection{Motivation}

@@

\subsubsection{Contribution}

@@

\subsubsection{Organization}

@@

\begin{itemize}
	\item {\bf Background:} Chapter~\ref{chap:background} will present the requisite background material on concurrent programming, stateless model checking, and the various types of programs targeted by Landslide.
	\item {\bf Landslide:} Chapter~\ref{chap:landslide} explains the design and implementation of Landslide
		%the core of my stateless model checking framework,
		and all the special features it's been equipped with over the years.
		It is the foundation upon which all three contributions above build.
	\item {\bf Quicksand:} Chapter~\ref{chap:quicksand} presents the Iterative Deepening framework which more intelligently chooses which state spaces to test, corresponding to contribution 1 above.
	\item {\bf Education:} Chapter~\ref{chap:education} discusses my evaluation of Landslide
		in CMU's 15-410 class environment using the Pebbles kernel,
		and in the University of Chicago's and Berkeley's OS class environments using the Pintos kernel,
		corresponding to contribution 2 above.
	%\item {\bf Pebbles:} Chapter~\ref{chap:410} discusses my evaluation of Landslide in CMU's 15-410 class environment using the Pebbles kernel, corresponding to contribution 2 above.
	%\item {\bf Pintos:} Chapter~\ref{chap:410} discusses my evaluation of Landslide in the University of Chicago's and Berkeley's OS class environments using the Pintos kernel, corresponding to contribution 2 above.
	\item {\bf Transactional Memory:} Chapter~\ref{chap:tm} presents my extension of Landslide's concurrency model to handle transactional concurrency and the evaluation thereof, corresponding to contribution 3 above.
	\item {\bf Related Work:} Chapter~\ref{chap:relatedwork} honors my neighbours and ancestors in research spirit.
	\item {\bf Future Work:} Chapter~\ref{chap:warpzone} lists stateless model checking's remaining shortcomings and how new research might address them.
	\item {\bf Conclusion:} Chapter~\ref{chap:conclusion} provides some thoughts on the future of the field.
\end{itemize}

@@

Pronoun use will vary between more specific and more ambiguous to convey additional nuance.
The singular ``I'' is associated with my own research contributions,
while the royal ``we'' should be taken to include the reader,
such as when surveying background material or related work,
to which the author and reader share more similar relationships.
The impersonal ``the programmer'' will be referred to as she/her
to challenge readers' unconscious bias about gender in computer science,
as well as to highlight her role as the intended user, separate from the underlying research.
%to distinguish her from the thesis audience concerned with the underlying research,
Individual students who participated in the user studies will be given the more inclusive they/them.
Gender-neutral pronouns will also be used on the author themself.

This document is, in a way, only half the work of the thesis, the other half being Landslide's implementation.
While some readers may prefer to be taught in prose and/or mathematical notation how an algorithm works,
others may find that disorienting and wish to see things in a way a compiler would understand.
Chapter~\ref{chap:landslide} provides source code links and serves as a guide to browsing the repository.
Later chapters will often make parenthetical references to specific files and/or functions
therein which implement a feature under discussion.

### 2018-10-06 thesis/landslide.tex
assorted minor prose revisions

Although a superhuman reader may quickly reach intuitive understanding
of complex algorithms from dense prose and mathematical notation alone,
I also present here a friendly example of using DPOR on Figure~\ref{fig:concurrency-bug}'s example program,
whose original state space is shown in Figure~\ref{fig:tree}.
In this program both threads are always enabled, imposing no scheduling constraints,
so memory conflicts alone will drive exploration.
%Acknowledging that quickly gaining an intuitive understanding for an algorithm
%is basically impossible from just paragraphs of text,
%I conclude this section with an example of a single iteration of DPOR
%on the example state space from Figures~\ref{fig:concurrency-bug} and \ref{fig:tree}.
First, let us consider a single iteration of DPOR, applied after executing the first branch.
The result is shown in Figure~\ref{fig:dpor-example-0}.

@@

Ultimately, the second half of the state space is pruned symmetrically.
In general, the number of ways to interleave two threads executing $N$ and $M$ events each is given by ${N+M \choose N}$%
\footnote{Generalizing to $K$ threads, and simplifying to the same number $N$ of events each,
this formula becomes $\frac{n!k!}{n!^k}$.};
in this case, the original state space's size was ${3+3 \choose 3} = 20$.
DPOR's reduction is characterized by replacing $N$ and $M$ with the number of {\em conflicting} events only;
in this case, ignoring all {\tt tmpN++} reorderings and testing only ${2+2 \choose 2} = 6$ branches.

@@

It is easy to see how these formulas might fit into DPOR's incremental search procedure:
at the end of each branch compute its $\mathsf{proportion}$ and add it in to a global $\mathsf{estimate}$ value.
However, DPOR may tag new branches to explore that would affect past branches' proportions,
requiring them to be recomputed,
which is not feasible without storing the entire exponentially-sized tree in memory.
Instead, Landslide also stores per-subtree estimates at intermediate $v_i$ nodes, $1 < i < n$,
along the current branch.
Whenever DPOR marks a new $k$th branch for exploration
at some $v_i$
%with $\mathsf{marked~children}(v_i) = k$,
its estimate
is multiplied by $(k-1)/k$ to retroactively adjust all past branches' proportions contributing to that estimate.
The change is also propagated to its sub-subtrees,
whose estimates must also incorporate the new $\mathsf{marked~children}$ value.
This allows Landslide to update the global estimate after each branch in $O(n)$ time and memory,
without recomputing past branch proportions individually.

\subsubsection{Runtime (Recursive) estimation}

>>>>

Although a superhuman reader may quickly reach intuitive understanding
of complex algorithms from dense prose and mathematical notation alone,
mortal readers may prefer the following example of using DPOR on the program from Figure~\ref{fig:concurrency-bug},
whose original state space is shown in Figure~\ref{fig:tree}.
In this program both threads are always enabled, imposing no scheduling constraints,
so memory conflicts alone will drive exploration.
%Acknowledging that quickly gaining an intuitive understanding for an algorithm
%is basically impossible from just paragraphs of text,
%I conclude this section with an example of a single iteration of DPOR
%on the example state space from Figures~\ref{fig:concurrency-bug} and \ref{fig:tree}.
First, let us consider a single iteration of DPOR, applied after executing the first branch.
The result is shown in Figure~\ref{fig:dpor-example-0}.

@@

Ultimately, the second half of the state space is pruned symmetrically.
In general, the number of ways to interleave two threads executing $N$ and $M$ events each is given by ${N+M \choose N}$%
\footnote{Generalizing to $K$ threads, and simplifying to $N$ events each,
this formula becomes $\frac{n!k!}{n!^k}$.};
in this case, the original state space's size was ${3+3 \choose 3} = 20$.
DPOR's reduction is characterized by replacing $N$ and $M$ with the number of {\em conflicting} events only;
in this case, ignoring all {\tt tmpN++} reorderings and testing only ${2+2 \choose 2} = 6$ branches.

@@

It is easy to see how these might fit into DPOR's incremental search procedure:
at the end of each branch compute its $\mathsf{proportion}$ and add it in to a global $\mathsf{estimate}$ value.
However, DPOR may tag new branches to explore that would affect past branches' proportions,
requiring them to be recomputed,
%which is not feasible without
which would require
storing the entire exponentially-sized tree in memory.
Instead, Landslide also stores per-subtree estimates at intermediate $v_i$ nodes, $1 < i < n$,
along the current branch.
Whenever DPOR marks a new $k$th branch
%for exploration
at some $v_i$
%with $\mathsf{marked~children}(v_i) = k$,
its estimate
is multiplied by $(k-1)/k$ to retroactively adjust all past branches' proportions contributing to that estimate.
The change is also propagated to its sub-subtrees,
whose estimates must also incorporate the new $\mathsf{marked~children}$ value.
This allows Landslide to update the global estimate after each branch in $O(n)$ time and memory,
without recomputing past branch proportions individually.

\subsubsection{Run-time (Recursive) estimation}

### 2018-10-06 thesis/overview.tex
assorted minor prose revisions

This chapter will introduce the necessary background material on concurrency, stateless model checking, data-race analysis, and the relevant undergraduate operating systems classes.

@@

		For example, {\em Landslide} heuristically judges the program to be stuck in an infinite loop

>>>>

This chapter introduces the necessary background material on concurrency, stateless model checking, data-race analysis, and the relevant undergraduate operating systems classes.

@@

		For another example, {\em Landslide} heuristically judges the program to be stuck in an infinite loop

### 2018-10-06 thesis/related.tex
assorted minor prose revisions

Equal partners in concurrency testing are the practical and the theoretical:
the former meaning tool implementations that target specific problem domains and help users as best one can,
and the latter meaning algorithmic advances to bring ever-larger state spaces within the realm of computational feasibility.
I discuss my most closely related works split in two sections accordingly.

@@

To date a number of techniques have been proposed to mitigate exponential explosion,
the Sisyphean rock of SCT.
The notion that some interleavings of concurrent threads could lead to indistinguishable program states and be therefore redundant,
known as {\em partial order reduction} (POR),
was first proposed in \cite{partial-model-checking}
and explored in detail in \cite{partial-order-methods}.
{\em Dynamic POR} (DPOR) was later developed in \cite{dpor},
proposing to track communication events between threads on-the-fly (i.e., dynamically)
rather than to rely on imprecise static alias analyses,
and is now considered the baseline for all subsequent state space reduction approaches in SCT.
That paper includes the {\em sleep sets} extension,
which Landslide includes in its implementation.
It is a {\em sound} reduction algorithm, meaning it will never fail to test a possible program behavior, despite skipping many execution sequences.
\sect{\ref{sec:landslide-dpor}} provides a detailed walk-through of how DPOR works,
as many of this thesis's contributions build directly upon it.

DPOR has since been extended in several ways to achieve further reduction
and to incorporate new concurrency models.
Optimal DPOR \cite{optimal-dpor} extends sleep sets into the more expressive {\em wakeup trees},
which provably tests exactly one interleaving from each equivalence class,
i.e., the optimal possible reduction,
at least under the memory independence definition of equivalence.
Extending the equivalence relation itself to capture not just memory {\em address} conflicts
but also the {\em values} read and written,
SATCheck \cite{satcheck} and Maximal Causality Reduction (MCR) \cite{mcr}
use an SMT solver \cite{z3} to identify additional pruning opportunities.
Implementing wakeup trees or SMT-driven exploration in Landslide is left to future work.

@@

seL4 \cite{sel4} is a microkernel fully designed and specified in Haskell and translated into C.
Its proofs guarantee not only standard security properties such as process isolation and bounded interrupt latency,
but also that the C code faithfully implements the specification.
It addresses concurrency by enabling system interrupts only at carefully-chosen code points,
and proving bounded runtime besides to ensure good preemptibility.
%and that the executable binary is correctly compiled from the C.
This degree of verification must however come at a cost:
seL4's authors reported over 2 person-years of development effort, with the majority spent on the Haskell specification.
%
CertiKOS \cite{certikos} extends this approach to include full concurrency and fine-grained locking
in the scope of verification,
using a proof in the Coq proof assistant
\cite{coq}
that also took 2 person-years to complete.
Its safety properties hold under all possible interleavings,
and include data-race freedom as well as standard sequential properties
such as no null dereference and no buffer or integer overflow,
although it stops short of reasoning about relaxed memory orderings or the TLB cache.
Many programmers would find a verification cost measured in person-years far too prohibitive,
while others might argue that for safety-critical kernel code you can't afford {\em not} to verify so thoroughly.
%
More recently, Hyperkernel \cite{hyperkernel} extended the xv6 educational kernel \cite{xv6}
to allow for partial, case-by-case verification of system call behaviour using state-machine specification in Python
checked by an SMT solver.
To limit verification complexity, it assumes not only uniprocessor execution
but also that interrupts be perpetually disabled,
taking concurrency entirely out of the equation to allow for greater extensibility
and lessen the programmer's verification burden.
%
Heroic as such end-to-end formal verification projects are,
this thesis finds that trading off thoroughness for accessibility is also acceptable
if it means helping more users overall.
Future work could check preemptive and/or multiprocessor kernels
%by combining Hyperkernel's approach,
by first checking safety properties in the absence of concurrency,
then checking with Landslide that concurrency introduces no new program behaviour not already verified,
to provide a less formal, but still hopefully useful, verification guarantee.
%
\revision{Lastly, CompCertTSO \cite{compcerttso} extends the CompCert verified compiler \cite{compcert}
to capture x86's relaxed memory semantics,
guaranteeing that a program written in the ClightTSO subset of C
is translated accurately to assembly with the same behaviour.
Like CertiKOS, its implementation is verified in Coq.
Although Landslide checks programs directly at the executable level,
blind to the code the programmer personally wrote,
extending that pipeline with a certifying compiler would
improve the overall verification,
ensuring that Landslide was checking the behaviour the programmer actually intended to write.}

>>>>

Equal partners in concurrency testing are the practical and the theoretical:
%the former meaning
tool implementations that target specific problem domains
%and manage practical tradeoffs,
%and help users as best one can,
%and the latter meaning
and algorithmic advances to make ever-larger state spaces computationally feasible.
%within the realm of computational feasibility.
The following two subsections discuss the most closely related prior work accordingly.
%I discuss my most closely related works split in two sections accordingly.

@@

To date a number of techniques have been proposed to mitigate exponential explosion,
the Sisyphean rock of SCT.
The notion that some interleavings of concurrent threads could lead to indistinguishable program states and be therefore redundant,
known as {\em partial order reduction} (POR),
was first proposed in \cite{partial-model-checking}
and explored in detail in \cite{partial-order-methods}.
{\em Dynamic POR} (DPOR) was later developed in \cite{dpor},
proposing to track communication events between threads on-the-fly (i.e., dynamically)
rather than to rely on imprecise static alias analyses,
and is now considered the baseline for all subsequent state space reduction approaches in SCT.
That paper includes the {\em sleep sets} extension,
which Landslide includes in its implementation.
It is a {\em sound} reduction algorithm, meaning it will never fail to test a possible program behavior, despite skipping many execution sequences.
\sect{\ref{sec:landslide-dpor}} provides a detailed walk-through of how DPOR works,
as many of this thesis's contributions build directly upon it.
%
DPOR has since been extended in several ways to achieve further reduction
and to incorporate new concurrency models.
Optimal DPOR \cite{optimal-dpor} extends sleep sets into the more expressive {\em wakeup trees},
which provably tests exactly one interleaving from each equivalence class,
i.e., the optimal possible reduction,
at least under the memory independence definition of equivalence.
Extending the equivalence relation itself to capture not just memory {\em address} conflicts
but also the {\em values} read and written,
SATCheck \cite{satcheck} and Maximal Causality Reduction (MCR) \cite{mcr}
use an SMT solver \cite{z3} to identify additional pruning opportunities.
Implementing wakeup trees or SMT-driven exploration in Landslide is left to future work.

@@

seL4 \cite{sel4} is a microkernel fully designed and specified in Haskell and translated into C.
Its proofs guarantee not only standard security properties such as process isolation and bounded interrupt latency,
but also that the C code faithfully implements the specification.
It addresses concurrency by enabling system interrupts only at carefully-chosen code points,
and proving bounded runtime besides to ensure good preemptibility.
%and that the executable binary is correctly compiled from the C.
This degree of verification must however come at a cost:
seL4's authors reported over 2 person-years of development effort, with the majority spent on the Haskell specification.
%
CertiKOS \cite{certikos} extends this approach to include full concurrency and fine-grained locking
in the scope of verification,
using a proof in the Coq proof assistant
\cite{coq}
that also took 2 person-years to complete.
Its safety properties hold under all possible interleavings,
and include data-race freedom as well as standard sequential properties
such as no null dereference and no buffer or integer overflow,
although it stops short of reasoning about relaxed memory orderings or the TLB cache.
Many programmers would find a verification cost measured in person-years far too prohibitive,
while others might argue that for safety-critical kernel code you can't afford {\em not} to verify so thoroughly.
%
More recently, Hyperkernel \cite{hyperkernel} extended the xv6 educational kernel \cite{xv6}
to allow for partial, case-by-case verification of system call behaviour using state-machine specification in Python
checked by an SMT solver.
To limit verification complexity, it assumes not only uniprocessor execution
but also that interrupts be perpetually disabled,
taking concurrency entirely out of the equation to allow for greater extensibility
and lessen the programmer's verification burden.
%
Heroic as such end-to-end formal verification projects are,
this thesis finds that trading off thoroughness for accessibility is also acceptable
if it means helping more users overall.
Future work could check preemptive and/or multiprocessor kernels
%by combining Hyperkernel's approach,
by first checking safety properties in the absence of concurrency,
then checking with Landslide that concurrency introduces no new program behaviour not already verified,
to provide a less formal, but still hopefully useful, verification guarantee.
%
\revision{Lastly, CompCertTSO \cite{compcerttso} extends the CompCert verified compiler \cite{compcert}
to capture x86's relaxed memory semantics,
guaranteeing that a program written in the ClightTSO subset of C
is translated accurately to assembly with the same behaviour.
Like CertiKOS, its implementation is verified in Coq.
Although Landslide checks programs directly at the executable level,
blind to the source code the programmer personally wrote,
extending that pipeline with a certifying compiler would
improve the overall verification,
ensuring that Landslide was actually checking the behaviour the programmer intended.}

### 2018-10-06 thesis/tm.tex
add missing section label in tm
assorted minor prose revisions

\subsubsection{Experimental setup}

@@

Note two curiosities:
firstly,
the impact that fixing Figure~\ref{fig:spinlockbug}'s performance bug
(changing a read+write to a read only)
had on even the correctness tests:
{\tt lock}({\tt spin\_fixed})'s state spaces were reduced by nearly half compared to {\tt lock}({\tt spinlock}),
on account of DPOR no longer needing to reorder the (now) read-read access pairs.
Secondly, {\tt spinlock} and {\tt spin\_fixed} take longer to test per interleaving
%(CPU time divided by state space size)
than {\tt mutex}
% TODO recalc
(roughly 9 interleavings per second for the former, 22 for the latter),
because while {\tt mutex} abstracts away threads needing to wait their turn for the critical section
behind an API Landslide understands,
the spinlock's wait loop is open-coded, and Landslide must fall back on its costlier heuristic synchronization detection
(\sect{\ref{sec:landslide-blocking}}).
In this way (and also, of course, because {\tt mutex}'s state spaces are smaller overall),
{\tt mutex} can be seen in turn as a further abstraction reduction of {\tt spinlock}.

@@

\section{Discussion}

@@

Likewise, the cases of {\tt htm2}(2,3) and (4,1),
in which retry sets achieved better reduction than STM mode,
show that the latter does not necessarily subsume the former,
and that combining the two could in theory achieve further reduction still.
However, {\tt xbegin} results other than {\tt \_XABORT\_RETRY}
may depend on the execution logic
(explicit aborts may be conditional on some change by a conflicting thread,
and conflict aborts cannot occur before their conflicting transaction to begin with%
\footnote{See {\tt 410user/progs/htm\_causality.c}; that was a fun realization to have
already halfway into implementing retry sets the wrong way at first.}%
),
so how to soundly prune either success paths or retry aborts
while other abort codes are in play
remains an open problem.

>>>>

\subsubsection{Experimental setup}
\label{sec:tm-eval-exp-setup}

@@

Note two curiosities:
firstly,
the impact that fixing Figure~\ref{fig:spinlockbug}'s performance bug
(changing a read+write to a read only)
had on even the correctness tests:
{\tt lock}({\tt spin\_fixed})'s state spaces were reduced by nearly half compared to {\tt lock}({\tt spinlock}),
on account of DPOR no longer needing to reorder the (now) read-read access pairs.
Secondly, {\tt spinlock} and {\tt spin\_fixed} take longer to test per interleaving
%(CPU time divided by state space size)
than {\tt mutex}
% old ver.
%(roughly 9 interleavings per second for the former, 22 for the latter),
% lock spin fixed:
% (4+998+138911+186+169893+286910) / (355+10557+3600000+2827+3600000+3600000) = 5.51
% lock mutex:
% (4+180+9372+489480+132+714882+15064) / (344+1683+40521+2499968+1500+3600000+66589) = 19.79
(roughly 6 interleavings per second for the former, 20 for the latter),
because while {\tt mutex} abstracts away threads needing to wait their turn for the critical section
behind an API Landslide understands,
the spinlock's wait loop is open-coded, and Landslide must fall back on its costlier heuristic synchronization detection
(\sect{\ref{sec:landslide-blocking}}).
In this way (and also, of course, because {\tt mutex}'s state spaces are smaller overall),
{\tt mutex} can be seen in turn as a further abstraction reduction of {\tt spinlock}.

@@

\section{Discussion}
\label{sec:tm-discussion}

@@

% TODO: is this true? at all? htm2 is bigger in STM only because it's got a separate expliit abort
Likewise, the cases of {\tt htm2}(2,3), (2,4), and (4,1),
in which retry sets achieved better reduction than STM mode,
show that the latter does not necessarily subsume the former,
and that combining the two could in theory achieve further reduction still.
However, {\tt xbegin} results other than {\tt \_XABORT\_RETRY}
may depend on the execution logic
(explicit aborts may be conditional on some change by a conflicting thread,
and conflict aborts cannot occur before their conflicting transaction to begin with%
\footnote{See {\tt 410user/progs/htm\_causality.c}; that was a fun realization to have
already halfway into implementing retry sets the wrong way at first.}%
),
so how to soundly prune either success paths or retry aborts
while other abort codes are in play
remains an open problem.
